Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars

### 2014-10-21 drafts/cfp/draft.tex
rework the discussion

\documentclass[11pt, twocolumn]{article}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{anyfontsize}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{titling}

@@

\title{Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars}
\author{Eyal Dechter,  Josh Rule, Josh Tenenbaum}

\begin{document}
\vspace{-10cm}
\maketitle

{\small For humans, learning abstract concepts and learning language
  go hand in hand: we acquire abstract knowledge primarily through
  linguistic experience, and acquiring abstract concepts is a crucial
  step in learning the meanings of linguistic expressions. Number
  knowledge is a case in point: we can only acquire concepts such as
  seventy-three through linguistic means, and we can only know what
  the sentence ``Seventy-three is more than twice as big as thirty-one"
  means if we can grasp the meanings of all its component number
  words. How, though, can machines learn systems of concepts from
  sentences about those concepts? One approach is to estimate the
  distribution from which those sentences are drawn, and, in doing so,
  infer the latent concepts and relationships that best explain the
  data. We present early work on a learning framework called Latent
  Predicate Networks (LPNs) which learns concepts by inferring the
  parameters of probabilistic context-sensitive grammars over
  sentences.  We show that for a small fragment of sentences
  expressing relationships between English number words, we can use
  hierarchical Bayesian inference to learn LPNs that can answer simple
  queries about previously unseen linguistic concepts within this
  domain. These generalizations demonstrate LPNs' promise as a tool
  for learning and representing conceptual knowledge in language.}

@@

As Claude Shannon demonstrated with his guessing game experiment in
1950~\cite{Shannon1950}, an adult English speaker's understanding of
her language and her world is such that the entropy of a character in
English text is slightly more than one bit. Plausibly, any machine
learning system able to predict natural language text with such
accuracy will have learned much about the concepts expressed in that
text. This intuition motivates us to invert Shannon's experiment and
learn linguistic concepts by learning a probabilistic generative model
for natural language.

@@

Our approach, then, will be to explore the use of probabilistic
context \emph{sensitive} grammars for modeling natural language,
extending the tools of hierarchical grammar induction~\cite{SOMEONE} from
context-free grammars to ones that can express not only syntactic
relationships but semantic ones as well. In the place of latent
syntactic categories, we will refer to latent predicates -- relations
between sentence components -- that, when learned, should correspond
to the concepts that underlie sentence semantics.

@@

\textbf{Example:} The following is an RCG for the 3-copy language
$\{www | w \in {a, b}^*\}$:
\begin{alignat*}{3}
&S(XYZ) &&\rightarrow A(X, Y, Z)\\
&A(aX, aY, aZ) &&\rightarrow A(X, Y, Z)\\
&A(bX, bY, bZ) &&\rightarrow A(X, Y, Z)\\
&A(\epsilon, \epsilon, \epsilon) &&\rightarrow 
\end{alignat*}

@@

LPNs can be encoded as a restricted and tractable subclass of
SLPs. Specifically, we encode LPNs with PRISM PROLOG to obtain
built-in predicates for probabilistic execution and Bayesian inference
over SLPs (Figure \ref{fig:copy}).

@@

@@

@@

\begin{figure}
	\rule{\linewidth}{3in}
	\caption{The Results}
	\label{fig:results}
\end{figure}

\begin{tabular}{l l}
Question & $N_p=4$ \\ \hline
 after twenty comes \underline{\hspace{1cm}}? & twenty one \checkmark \\
 after forty five comes \underline{\hspace{1cm}}? & forty six \checkmark \\
 after forty seven comes \underline{\hspace{1cm}}? & forty eight  \checkmark \\
 after forty nine comes \underline{\hspace{1cm}}? & forty ten $\times$ \\
 after fifty nine comes \underline{\hspace{1cm}}? & fifty ten $\times$ \\
 after sixty one comes \underline{\hspace{1cm}}? & sixty two \checkmark \\
 after sixty three comes \underline{\hspace{1cm}}? & sixty four \checkmark \\
 after sixty four comes \underline{\hspace{1cm}}? & sixty five \checkmark \\
 after sixty five comes \underline{\hspace{1cm}}? & sixty six \checkmark \\
 after sixty nine comes \underline{\hspace{1cm}}? & sixty ten \checkmark \\
 after seventy three comes \underline{\hspace{1cm}}? & seventy four \checkmark \\
 after seventy nine comes \underline{\hspace{1cm}}? & seventy ten $\times$ \\
 after ninety five comes \underline{\hspace{1cm}}? & ninety six \checkmark \\
 before twenty three comes \underline{\hspace{1cm}}? & twenty two \checkmark \\
 before thirty comes \underline{\hspace{1cm}}? & thirty eighty $\times$ \\
 before thirty eight comes \underline{\hspace{1cm}}? & thirty seven \checkmark \\
 before forty one comes \underline{\hspace{1cm}}? & forty \checkmark \\
 before fifty three comes \underline{\hspace{1cm}}? & fifty two \checkmark \\
 before sixty eight comes \underline{\hspace{1cm}}? & sixty seven \checkmark \\
 before seventy two comes \underline{\hspace{1cm}}? & seventy one \checkmark \\
 before seventy three comes \underline{\hspace{1cm}}? & seventy two \checkmark \\
 before eighty five comes \underline{\hspace{1cm}}? & eighty four \checkmark \\
 before ninety two comes \underline{\hspace{1cm}}? & ninety one \checkmark \\
 before ninety three comes \underline{\hspace{1cm}}? & ninety two \checkmark \\
 before ninety five comes \underline{\hspace{1cm}}? &ninety four \checkmark \\

\end{tabular}

\begin{alignat*}{2}
&S(X Y) \leftarrow A_1(X, Y) : 1.0000 \\
&A_1(X Y, U V) \leftarrow A_2(X, U), A_3(V, Y) : 0.5002 \\
&A_2(before, comes) : 0.7316 \\
&A_4(after, comes) : 0.9990 \\
&A_1(X Y, U V) \leftarrow A_3(Y, V), A_4(X, U) : 0.3428 \\
&A_3(one, two) : 0.3993 \\
&A_3(two, three) : 0.2063 \\
&A_3(three, four) : 0.1093 \\
&A_1(X Y, U V) \leftarrow A_1(V, Y), A_4(X, U) : 0.0796 \\
&A_1(X Y, U V) \leftarrow A_1(Y, V), A_2(X, U) : 0.0712 \\
&A_3(four, five) : 0.0734 \\
&A_3(five, six) : 0.0502 \\
&A_3(six, seven) : 0.0355 \\
&A_3(eight, nine) : 0.0290 \\
&A_3(seven, eight) : 0.0271 \\
&A_2(fifty, fifty) : 0.0375 \\
&A_2(thirty, thirty) : 0.0361 \\
&A_2(eighty, eighty) : 0.0339 \\
&A_3(null, one) : 0.0231 \\
&A_2(forty, forty) : 0.0332 \\
&A_2(twenty, twenty) : 0.0310 \\
&A_2(seventy, seventy) : 0.0296 \\
&A_2(sixty, sixty) : 0.0274 \\
&A_2(ninety, ninety) : 0.0260 \\
&A_3(eighteen, nineteen) : 0.0064 \\
&A_3(sixteen, seventeen) : 0.0049 \\
&A_3(eleven, twelve) : 0.0044 \\
&A_3(nine, ten) : 0.0044 \\
&A_3(thirteen, fourteen) : 0.0044 \\
&A_3(fourteen, fifteen) : 0.0039 \\
&A_3(ten, eleven) : 0.0034 \\
&A_2(null, fifty) : 0.0043 \\
&A_3(eighty, null) : 0.0030 \\
&A_3(seventeen, eighteen) : 0.0030 \\
&A_1(X Y, U V) \leftarrow A_2(Y, X), A_3(V, U) : 0.0021 \\
&A_3(nine, sixty) : 0.0025 \\
&A_2(nine, seventy) : 0.0029 \\
&A_3(nine, forty) : 0.0020 \\
&A_1(X Y, U V) \leftarrow A_1(V, Y), A_2(X, U) : 0.0013 \\
&A_2(null, thirty) : 0.0022 \\
&A_2(nine, ninety) : 0.0022 \\
&A_3(twelve, thirteen) : 0.0015 \\
&A_3(fifteen, sixteen) : 0.0015 \\
&A_1(X Y, U V) \leftarrow A_2(V, X), A_3(Y, U) : 0.0008 \\
&A_3(null, comes) : 0.0010 \\
&A_1(X Y, U V) \leftarrow A_1(X, U), A_2(V, Y) : 0.0008 \\
&A_1(X Y, U V) \leftarrow A_1(X, U), A_2(Y, V) : 0.0008 \\
&A_2(twenty, after) : 0.0014 \\
&A_2(sixty, before) : 0.0007 \\
&A_3(nineteen, comes) : 0.0005 \\
&A_4(nine, thirty) : 0.0010 \\
&A_1(X Y, U V) \leftarrow A_1(X, U), A_4(Y, V) : 0.0004 \\
\end{alignat*} \\
\section{Discussion}

LPNs are only one of many models aimed at jointly learning the
structure and meaning of natural language expressions, with semantic
parsers being a popular alternative. Semantic parsers, however, learn
separate systems of syntax and semantics, often pairing Combinatory
Categorial Grammars of syntax with an explicit logical language of
semantics, typically the untyped lambda
calculus~\cite{berant2013semantic, liang2013learning,
  kwiatkowski2010inducing}. By contrast, our approach uses a single
formalism, the LPN, to describe both which sentences are syntactically
valid and what sorts of semantic information those sentences
imply. Semantic parsers also typically learn models of semantics by
pairing sentences with an explicit logical form or real-world
denotation. In our model it is the grammar induction over sentences
alone which reveals candidate relations between words that are key to
capturing their meaning. Future work should seek to understand the
limits of modeling syntactic and semantic information as inseparable
parts of as a single system, and whether there might be ways to
flexibly combine the strengths of this approach with more traditional
dual system models.

Previous work has shown that generative models of probabilistic
programs are a useful tool for understanding human concept learning
for domains as varied as handwritten characters and structured visual
objects~\cite{dechter2013bootstrap, Stuhlmuller, BrendensPaper}. Here,
we continue that work by showing how some forms of both human language
learning and concept learning may be fruitfully modeled as inference
over a generative model of probabilistic logic programs. Future work
should work explore how LPNs and its extensions may apply to other
domains where generative models of probabilistic programming have
already proven useful, as well as generalizing the current approach to
linguistic concepts.

>>>>

\documentclass[10pt, twocolumn]{article}

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{anyfontsize}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{titling}
\usepackage{array}
%\usepackage{multicol}

@@

\title{Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars}
\author{Eyal Dechter,  Josh Rule, Josh Tenenbaum}
\date{}

\begin{document}
\vspace{-10cm}
\maketitle
\vspace{-2em}

{\small For humans, learning abstract concepts and learning language
  go hand in hand: we acquire abstract knowledge primarily through
  linguistic experience, and acquiring abstract concepts is a crucial
  step in learning the meanings of linguistic expressions. Number
  knowledge is a case in point: we largely acquire concepts such as
  seventy-three through linguistic means, and we can only know what
  the sentence ``Seventy-three is more than twice as big as
  thirty-one" means if we can grasp the meanings of its component
  number words. How do we make sense of this potential paradox? One
  approach is to estimate the distribution from which sentences
  are drawn, and, in doing so, infer the latent concepts and
  relationships that best explain those sentences. We present early work on a
  learning framework called Latent Predicate Networks (LPNs) which
  learns concepts by inferring the parameters of probabilistic
  context-sensitive grammars over sentences.  We show that for a small
  fragment of sentences expressing relationships between English
  number words, we can use hierarchical Bayesian inference to learn
  LPNs that can answer simple queries about previously unseen
  abstract concepts within this domain. These generalizations
  demonstrate LPNs' promise as a tool for learning and representing
  conceptual knowledge in language.}

@@

As Claude Shannon demonstrated with his guessing game experiment in
1950~\cite{shannon1951prediction}, an adult English speaker's understanding of
language and the world is such that the entropy of a character in
English text is slightly more than one bit. Plausibly, any machine
learning system able to predict natural language text with human-like
accuracy will have learned much about the concepts expressed in that
text. This intuition motivates us to invert Shannon's experiment and
learn linguistic concepts by learning a probabilistic generative model
for natural language.

@@

Our approach, then, will be to explore the use of probabilistic
context \emph{sensitive} grammars for modeling natural language,
extending the tools of hierarchical grammar induction~\cite{johnson2006adaptor, goldwater2006contextual, DBLP:conf/emnlp/LiangPJK07} from
context-free grammars to ones that can express not only syntactic
relationships but semantic ones as well. In the place of latent
syntactic categories, we will refer to latent predicates -- relations
between sentence components -- that, when learned, should correspond
to the concepts that underlie sentence semantics.

@@

\textbf{Example:} The following is an RCG for the 3-copy language
$\{www | w \in \{a, b\}^*\}$:
\begin{alignat*}{3}
&S(XYZ) &&\rightarrow A(X, Y, Z)\\
&A(aX, aY, aZ) &&\rightarrow A(X, Y, Z)\\
&A(bX, bY, bZ) &&\rightarrow A(X, Y, Z)\\
&A(\epsilon, \epsilon, \epsilon) &&\rightarrow 
\end{alignat*}

@@

LPNs can be encoded as a restricted and tractable subclass of
SLPs. Specifically, we encode LPNs with PRISM to obtain built-in
predicates for probabilistic execution and Bayesian inference over
SLPs (Figure \ref{fig:copy}). PRISM is an appropriate choice here,
because it implements a previously related extension of the
variational Bayesian Expectation Maximization algorithm for PCFGs to
SLPs satisfying a number of restrictions
\cite{cussens2001parameter}. Our LPN encodings satisfy these
restrictions, in particular the need for a finite explanation graph
({\it i.e.} set of all possible parse trees) for any
utterance. Because we disallow $A_i([\,],[\,])$ as a valid clause,
every term in the body of a clause has shorter arguments than the
head, giving finite derivations and a finite explanation graph.

@@

\begin{table}[t]
\begin{tabular}{>{\footnotesize} l >{\footnotesize} l}
  Question & $K=4$ \\ \hline
  after twenty comes \underline{\hspace{1cm}}? & twenty one \checkmark \\
  after forty five comes \underline{\hspace{1cm}}? & forty six \checkmark \\
  after forty seven comes \underline{\hspace{1cm}}? & forty eight  \checkmark \\
  after forty nine comes \underline{\hspace{1cm}}? & forty ten $\times$ \\
  after fifty nine comes \underline{\hspace{1cm}}? & fifty ten $\times$ \\
  after sixty one comes \underline{\hspace{1cm}}? & sixty two \checkmark \\
  after sixty three comes \underline{\hspace{1cm}}? & sixty four \checkmark \\
  after sixty four comes \underline{\hspace{1cm}}? & sixty five \checkmark \\
  after sixty five comes \underline{\hspace{1cm}}? & sixty six \checkmark \\
  after sixty nine comes \underline{\hspace{1cm}}? & sixty ten $\times$ \\
  after seventy three comes \underline{\hspace{1cm}}? & seventy four \checkmark \\
  after seventy nine comes \underline{\hspace{1cm}}? & seventy ten $\times$ \\
  after ninety five comes \underline{\hspace{1cm}}? & ninety six \checkmark \\
  before twenty three comes \underline{\hspace{1cm}}? & twenty two \checkmark \\
  before thirty comes \underline{\hspace{1cm}}? & thirty eighty $\times$ \\
  before thirty eight comes \underline{\hspace{1cm}}? & thirty seven \checkmark \\
  before forty one comes \underline{\hspace{1cm}}? & forty \checkmark \\
  before fifty three comes \underline{\hspace{1cm}}? & fifty two \checkmark \\
  before sixty eight comes \underline{\hspace{1cm}}? & sixty seven \checkmark \\
  before seventy two comes \underline{\hspace{1cm}}? & seventy one \checkmark \\
  before seventy three comes \underline{\hspace{1cm}}? & seventy two \checkmark \\
  before eighty five comes \underline{\hspace{1cm}}? & eighty four \checkmark \\
  before ninety two comes \underline{\hspace{1cm}}? & ninety one \checkmark \\
  before ninety three comes \underline{\hspace{1cm}}? & ninety two \checkmark \\
  before ninety five comes \underline{\hspace{1cm}}? &ninety four \checkmark \\
\end{tabular}
\caption{results of learning with 4 predicates and XXX}
\label{tab:results}
\end{table}

@@

{\it Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec a
  diam lectus. Sed sit amet ipsum mauris. Maecenas congue ligula ac
  quam viverra nec consectetur ante hendrerit. Donec et mollis
  dolor. Praesent et diam eget libero egestas mattis sit amet vitae
  augue. Nam tincidunt congue enim, ut porta lorem lacinia
  consectetur. Donec ut libero sed arcu vehicula ultricies a non
  tortor. Lorem ipsum dolor sit amet, consectetur adipiscing
  elit. Aenean ut gravida lorem. Ut turpis felis, pulvinar a semper
  sed, adipiscing id dolor. Pellentesque auctor nisi id magna
  consequat sagittis. Curabitur dapibus enim sit amet elit pharetra
  tincidunt feugiat nisl imperdiet. Ut convallis libero in urna
  ultrices accumsan. Donec sed odio eros. Donec viverra mi quis quam
  pulvinar at malesuada arcu rhoncus. Cum sociis natoque penatibus et
  magnis dis parturient montes, nascetur ridiculus mus. In rutrum
  accumsan ultricies. Mauris vitae nisi at sem facilisis semper ac in
  est.

  Vivamus fermentum semper porta. Nunc diam velit, adipiscing ut
  tristique vitae, sagittis vel odio. Maecenas convallis ullamcorper
  ultricies. Curabitur ornare, ligula semper consectetur sagittis,
  nisi diam iaculis velit, id fringilla sem nunc vel mi. Nam dictum,
  odio nec pretium volutpat, arcu ante placerat erat, non tristique
  elit urna et turpis. Quisque mi metus, ornare sit amet fermentum et,
  tincidunt et orci. Fusce eget orci a orci congue vestibulum. Ut
  dolor diam, elementum et vestibulum eu, porttitor vel
  elit. Curabitur venenatis pulvinar tellus gravida ornare. Sed et.}

@@

\begin{table}[t]
\begin{tabular}{>{\tiny} l >{\tiny} l}
\multicolumn{2}{>{\tiny}c}{$S$(X Y) $\leftarrow$ $A_1$(X, Y) : 1.0000} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(X, U), $A_3$(V, Y) : 0.5002} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_3$(Y, V), $A_4$(X, U) : 0.3428} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(V, Y), $A_4$(X, U) : 0.0796} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(Y, V), $A_2$(X, U) : 0.0712} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(Y, X), $A_3$(V, U) : 0.0021} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(V, Y), $A_2$(X, U) : 0.0013} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(V, X), $A_3$(Y, U) : 0.0008} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_2$(V, Y) : 0.0008} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_2$(Y, V) : 0.0008} \\
\multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_4$(Y, V) : 0.0004} \\
$A_2$(before, comes) : 0.7316 & $A_4$(after, comes) : 0.9990  \\
$A_3$(one, two) : 0.3993 &
$A_3$(two, three) : 0.2063 \\
$A_3$(three, four) : 0.1093 &
$A_3$(four, five) : 0.0734 \\
$A_3$(five, six) : 0.0502 &
$A_3$(six, seven) : 0.0355 \\
$A_3$(eight, nine) : 0.0290 &
$A_3$(seven, eight) : 0.0271 \\
$A_2$(fifty, fifty) : 0.0375 &
$A_2$(thirty, thirty) : 0.0361 \\
$A_2$(eighty, eighty) : 0.0339 &
$A_3$(null, one) : 0.0231 \\
$A_2$(forty, forty) : 0.0332 &
$A_2$(twenty, twenty) : 0.0310 \\
$A_2$(seventy, seventy) : 0.0296 &
$A_2$(sixty, sixty) : 0.0274 \\
$A_2$(ninety, ninety) : 0.0260 &
$A_3$(eighteen, nineteen) : 0.0064 \\
$A_3$(sixteen, seventeen) : 0.0049 &
$A_3$(eleven, twelve) : 0.0044 \\
$A_3$(nine, ten) : 0.0044 &
$A_3$(thirteen, fourteen) : 0.0044 \\
$A_3$(fourteen, fifteen) : 0.0039 &
$A_3$(ten, eleven) : 0.0034 \\
$A_2$(null, fifty) : 0.0043 &
$A_3$(eighty, null) : 0.0030 \\
$A_3$(seventeen, eighteen) : 0.0030 &
$A_3$(nine, sixty) : 0.0025 \\
$A_2$(nine, seventy) : 0.0029 &
$A_3$(nine, forty) : 0.0020 \\
$A_2$(null, thirty) : 0.0022 &
$A_2$(nine, ninety) : 0.0022 \\
$A_3$(twelve, thirteen) : 0.0015 &
$A_3$(fifteen, sixteen) : 0.0015 \\
$A_3$(null, comes) : 0.0010 &
$A_2$(twenty, after) : 0.0014 \\
$A_2$(sixty, before) : 0.0007 &
$A_3$(nineteen, comes) : 0.0005 \\
$A_4$(nine, thirty) : 0.0010 & \\
\end{tabular}
\caption{XXX}
\label{tab:grammar}
\end{table}

\section{Discussion}

LPNs seems particularly well-suited to learning in domains where the
number of concepts is infinite. In domains like these, the only way to
understand each concept is to discover structure in the sentences that
refer to these concepts. While number is a prime example, others
include spatial relations, kinship and other social relations, and
temporal relations. Future work should explore how LPNs might explain
learning in these domains, as well as how more complex number
relations might be learned.

Many of the algorithms and formalisms used here were originally
developed for use in other closely related areas, including logic
programming and semantic parsing. Semantic parsing, in particular,
seems especially related to the challenge we face here of jointly
learning meaning and structure from
sentences~\cite{berant2013semantic, liang2013learning,
  kwiatkowski2010inducing}. Semantic parsing, however, tends to frame
this challenge slightly differently than we do here, namely, by asking
how utterances can be mapped to an explicit internal logical
language. By contrast, we focus here on systems where meaning and
structure seem inseparable. Understanding how these two approaches
relate and inform one another is an interesting and open question.

%% Previous work has shown that generative models of probabilistic
%% programs are a useful tool for understanding human concept learning
%% for domains as varied as handwritten characters and structured visual
%% objects~\cite{dechter2013bootstrap, stuhlmuller2010learning,
%%   lake2012concept}. Here, we continue that work by showing how some
%% forms of both human language learning and concept learning may be
%% fruitfully modeled as inference over a generative model of
%% probabilistic logic programs. Future work should work explore how LPNs
%% and its extensions may apply to other domains where generative models
%% of probabilistic programming have already proven useful, as well as
%% generalizing the current approach to abstract, symbolic concepts.
